%-------------------------------------------------------------------------------
% Preamble
%-------------------------------------------------------------------------------

\documentclass[a4paper, 12pt]{article}
%\usepackage{ms}            % load the template
\usepackage[osf]{mathpazo} % palatino
%\usepackage[round]{natbib} % author-year citations
\usepackage[superscript,biblabel,nomove]{cite} % for superscript citations
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{parskip} 
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{pdflscape}

\pagenumbering{arabic}  
\linespread{1.66}

\begin{document}
%-------------------------------------------------------------------------------
% Title page information
%-------------------------------------------------------------------------------

\section*{Microchromosome fusions underpin convergent evolution of chameleon karyotypes}

\textbf{Short running title:} Chameleon karyotype evolution\\

\begin{centering}

Marcello Mezzasalma$^{1,2}$, 
Jeffrey W. Streicher$^{1}$, 
Fabio M. Guarino$^{3}$,\\ 
Marc E.H. Jones$^{1,4,5}$, 
Simon P. Loader$^{1}$, 
Gaetano Odierna$^{3}$,\\ 
and Natalie Cooper$^{1*}$

\end{centering}

  $^1$Science Group, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.\\
  $^2$Department of Biology, Ecology and Earth Science, University of Calabria, Via P. Bucci 4/B, 87036 Rende, Italy.\\
  $^3$Department of Biology, University of Naples Federico II, Via Cinthia 26, 80126 Naples, Italy.\\
  $^4$Research Department of Cell and Developmental Biology, Gower Street, UCL, University College London, London, UK.\\ 
  $^5$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia, SA5005, Australia.\\
  $*$Email address: natalie.cooper@nhm.ac.uk\\


\section*{Author contributions}
MM, JWS, MEHJ and SPL and NC conceptualised the study. MM collated data and performed molecular and cytogenetic analyses. MM, GO and FMG critically evaluated the cytogenetic results. NC ran the statistical analyses. MM, JWS and NC wrote the first draft. All authors reviewed and edited drafts and approved the final version for publication.

\section*{Acknowledgments}
We thank Gennaro Aprea, Frank Glaw and Franco Andreone for providing us the tissue samples and cell suspensions, and Malagasy authorities for granting research and export permits. We thank Tony Gamble and Peter Olson for discussions on chromosome evolution and Angelica Crottini for taxonomic insights.

\section*{Data accessibility}\label{data-code-and-materials}
All of the data are available on the NHM Data Portal\cite{chameleon-data}. 
R code to reproduce our analyses is available at github.com/nhcooper123/chameleon-chromosomes (Zenodo DOI will be added on acceptance\cite{coopercode2022}). 
The sequences generated in this study are available from GenBank with the primary accession codes XXX to be added on acceptance.

\section*{Funding}
MM was funded by the European Union’s Horizon 2020 research and innovation programme under the Marie Sk\l{}odowska-Curie grant agreement No 797027.

\section{Competing interests}
We declare we have no competing interests.

\bibliographystyle{bibstyle}
\bibliography{chameleons}

\end{document}