---
title: "Preparing the data for chromosome evolution models"
author: "Natalie Cooper"
output: html_document
---

```{r, message = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(ape)
```

## Prepare data and trees for ChromEvol analyses

Read in the data and extract species names and n
```{r, message = FALSE}
# Read in the data
ds <- read_csv(here("data/chromosome-data.csv"))

# Exclude Leiolepsis
ds <- ds %>% 
  filter(tips != "Leiolep")

# Exclude Rieppeleon
ds_exclude <- ds %>% 
  filter(tips != "Riebrev")

# Retain only species names and haploid chromosome numbers
mysample <- ds %>%
  select(tips, haploidn)

mysample_exclude <- ds_exclude %>%
  select(tips, haploidn)

# Look at it
# mysample
# mysample_exclude
```

Read in the tree, check it is ultrametric and remove node labels
```{r}
# Read in the tree
tree <- read.tree(here("data/ML-tree.tre"))
# Tree is not ultrametric due to weird rounding issues
# Fix this
tree <- force.ultrametric(tree)
# Check tree is now ultrametric
is.ultrametric(tree) # TRUE
# Remove node labels
tree$node.label <- NULL
# Remove "Leiolep" from the tree
tree <- drop.tip(tree, "Leiolep")
# Remove "Riebrev" from the tree
tree_exclude <- drop.tip(tree, "Riebrev")

# Look at it
# plot(tree, cex = 0.6)
# plot(tree_exclude, cex = 0.6)
```

Change the data structure so it can be used in ChromEvol
```{r}
# ChromEvol requires data in this format (where numbers are chromosome counts as haploid number):
# >SpeciesA
# 14
# >SpeciesB
# 28

# Add > in front of each species name
chromo <- mysample %>%
  mutate(tips = paste0(">", tips))

chromo_exclude <- mysample_exclude %>%
  mutate(tips = paste0(">", tips))

# Transpose the data so it is all in one column
chromo <- as.vector(t(chromo))
chromo_exclude <- as.vector(t(chromo_exclude))

# Make the vectore back into a dataframe
chromo <- as.data.frame(chromo)
chromo_exclude <- as.data.frame(chromo_exclude)

# Check it
#chromo
#chromo_exclude
```

Save the tree and data for use in ChromEvol
```{r}
# Save tree
write.tree(tree, here("chromEvol/tree_for_chromevol"))
write.tree(tree_exclude, here("chromEvol/tree_for_chromevol_exclude"))
# Save data
write_csv(chromo, col_names =  FALSE, 
          path = here("chromEvol/data_for_chromevol"))
write_csv(chromo_exclude, col_names =  FALSE, 
          path = here("chromEvol/data_for_chromevol_exclude"))
```

> Now go to ChromEvol package in the terminal and run it...

## For models at species level

```{r}
# Exclude duplicated species
ds_species <-
  ds %>% 
  mutate(dup = duplicated(ds$Binomial)) %>%
  filter(dup != TRUE)

ds_exclude_species <-
  ds_exclude %>% 
  mutate(dup = duplicated(ds_exclude$Binomial)) %>%
  filter(dup != TRUE)

# Retain only species names and haploid chromosome numbers
mysample <- ds_species %>%
  select(tips, haploidn)

mysample_exclude <- ds_exclude_species %>%
  select(tips, haploidn)

# Look at it
# mysample
# mysample_exclude
```

Remove excess species from tree
```{r}
# Remove species from the tree that are not in ds_species
remove <- setdiff(tree$tip.label, ds_species$tips)
speciestree <- drop.tip(tree, remove)
speciestree_exclude <- drop.tip(tree_exclude, remove)
```

Change the data structure so it can be used in ChromEvol
```{r}
# ChromEvol requires data in this format (where numbers are chromosome counts as haploid number):
# >SpeciesA
# 14
# >SpeciesB
# 28

# Add > in front of each species name
chromo <- mysample %>%
  mutate(tips = paste0(">", tips))

chromo_exclude <- mysample_exclude %>%
  mutate(tips = paste0(">", tips))

# Transpose the data so it is all in one column
chromo <- as.vector(t(chromo))
chromo_exclude <- as.vector(t(chromo_exclude))

# Make the vectore back into a dataframe
chromo <- as.data.frame(chromo)
chromo_exclude <- as.data.frame(chromo_exclude)

# Check it
#chromo
#chromo_exclude
```

Save the tree and data for use in ChromEvol
```{r}
# Save tree
write.tree(speciestree, here("chromEvol/tree_for_chromevol_species"))
write.tree(speciestree_exclude, here("chromEvol/tree_for_chromevol_exclude_species"))
# Save data
write_csv(chromo, col_names =  FALSE, 
          path = here("chromEvol/data_for_chromevol_species"))
write_csv(chromo_exclude, col_names =  FALSE, 
          path = here("chromEvol/data_for_chromevol_exclude_species"))
```

> Now go to ChromEvol package in the terminal and run it...