---
title: "Extracting the outputs of ChromEvol - OTUs"
author: "Natalie Cooper"
output: html_document
---

## Extracting outputs from ChromEvol

```{r, message = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(phytools)
library(ChromEvol) #install this from source
library(here)
```

### Model comparison
First look at the models and choose the best using AIC, and also extract the model parameters. 

Note that due to the rescaling of the tree, we need to multiply these values by the scaling constant to get the true values. This is specified in the `chromEvol.res` outfile, but doesn't seem to be automatically extractable so I've done this by hand to create the vector below. Note that it refers to the tree, so there are only two repeated values, one for the taxon tree with Riepelleon, one for the taxon tree without Riepelleon.

```{r}
rescale_param <- rep(c(0.00888284, 0.00826847), 3)
```

Now loop through the 6 sets of models
```{r, message = FALSE}
# Get a list of the output folder names
output_names <- list.files(path = here("chromEvol/OTUs/results"), pattern = "_out")
output <- NULL

# Loop through all 6 folders of outputs
for (i in 1:6) {
  # Read in the model summary data
  summary <- read_delim(here(paste0("chromEvol/OTUs/results/", output_names[i], "/models_summary.txt")), 
                        delim = "\t")

  # Select only the two models we want
  summary <-
    summary %>%
    filter(MODEL == "CONST_RATE_NO_DUPL" | MODEL == "LINEAR_RATE_NO_DUPL") %>%
    # Add model set name
    mutate(name = output_names[i]) %>%
    # reorder
    dplyr::select(name, everything())

  # Add parameter values
  # Read in the parameter results by specifying the path to the folder for each model
  # This throws a lot of warnings about NAs because of the strange ways these outputs are
  # contructed. But it works fine
  resultsC <- read.ce(here(paste0("chromEvol/OTUs/results/", output_names[i], "/CONST_RATE_NO_DUPL")))
  resultsL <- read.ce(here(paste0("chromEvol/OTUs/results/", output_names[i], "/LINEAR_RATE_NO_DUPL")))
  
  # Add the parameters to the summary
  # Note these will end up doubled up due to the results here being for each model
  # and the results in the summary having a line per model
  summary <-
    summary %>%
    mutate(CR_lossC = resultsC$Parameters$rate_lossC * rescale_param[i]) %>%
    mutate(CR_gainC = resultsC$Parameters$rate_gainC * rescale_param[i]) %>%
    mutate(LR_lossC = resultsL$Parameters$rate_lossC * rescale_param[i]) %>%
    mutate(LR_gainC = resultsL$Parameters$rate_gainC * rescale_param[i]) %>%
    mutate(LR_lossL = resultsL$Parameters$rate_lossL * rescale_param[i]) %>%
    mutate(LR_gainL = resultsL$Parameters$rate_gainL * rescale_param[i])  
  
  output <- rbind(output, summary)
}

# Look and save at the outputs
output
#write_csv(output, file = here("outputs/OTUs/chromEvol-model-summaries.csv"))
```