---
title: "Collating simulations of chromosome numbers"
author: "Natalie Cooper"
output: html_document
---

## Collate simulations results and create null simulations

```{r, message = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(phytools)
library(ChromEvol) #install this from source
library(patchwork)
```

Read in all the simulations outputs, wrangle into a useable format, and combine.
```{r, message = FALSE, warning = FALSE}
sim_output <- data.frame(haploidn = NULL, tips = NULL, sim = NULL)

# Loop through all 1000 folders of outputs
for (i in 0:999) {
  # Choose directory
  dir <- paste0("chromEvol/chromEvol_out_simulations/", i, "/")
  # Read in counts data
  counts <- read_delim(here(paste0(dir,"simCounts.txt")), 
                       delim = "\t", col_names = FALSE)

  # wrangle and add sim number
  countdata <- 
  counts %>%
  mutate(ind = rep(c("tips","haploidn"), length.out = n())) %>%
  group_by(ind) %>%
  mutate(id = row_number()) %>%
  spread(ind, X1) %>%
  dplyr::select(-id) %>%
  mutate(tips = gsub(">", "", tips)) %>%
  mutate(sim = rep(i+1, length(tips)))
  
  # add to output
  sim_output <- rbind(sim_output, countdata)
}

sim_output$sim <- sim_output$sim+1

write_csv(sim_output, path = here("data/chromevol-simulations-outputs.csv"))
```

## Null model simulations

Create null simulations by randomly assigning chromosome numbers to the tips. The numbers come from the data and sampling is done without replacement.

```{r}
# Read in observed data
mydata <- read_csv(here("data/chromosome-data.csv"))

# Remove Rieppeleon
mydata <- 
  mydata %>% 
  filter(genus != "Rieppeleon")

# Rename the haploidn column
# Select only tip names, genus and observed haploidn
mydata <-
  mydata %>%
  rename(obs_haploidn = haploidn) %>%
  dplyr::select(tips, obs_haploidn, genus)
```

Run simulations
```{r}
sims_output_null <- data.frame(sim_haploidn = NULL, tips = NULL, sim = NULL)
  
for (j in 1:1000){
  newdata <- data.frame(sim_haploidn = sample(mydata$obs_haploidn, 
                                              length(mydata$tips)),
                        tips = mydata$tips, 
                        sim = rep(j, length(mydata$tips)))

  sims_output_null <- rbind(sims_output_null, newdata)
}

write_csv(sims_output_null, path = here("data/null-simulations-outputs.csv"))
```

## SPECIE LEVEL

## Collate simulations results and create null simulations

Read in all the simulations outputs, wrangle into a useable format, and combine.
```{r, message = FALSE, warning = FALSE}
sim_output_species <- data.frame(haploidn = NULL, tips = NULL, sim = NULL)

# Loop through all 1000 folders of outputs
for (i in 0:999) {
  # Choose directory
  dir <- paste0("chromEvol/chromEvol_out_simulations_species/", i, "/")
  # Read in counts data
  counts <- read_delim(here(paste0(dir,"simCounts.txt")), 
                       delim = "\t", col_names = FALSE)

  # wrangle and add sim number
  countdata <- 
  counts %>%
  mutate(ind = rep(c("tips","haploidn"), length.out = n())) %>%
  group_by(ind) %>%
  mutate(id = row_number()) %>%
  spread(ind, X1) %>%
  dplyr::select(-id) %>%
  mutate(tips = gsub(">", "", tips)) %>%
  mutate(sim = rep(i+1, length(tips)))
  
  # add to output
  sim_output_species <- rbind(sim_output_species, countdata)
}

sim_output_species$sim <- sim_output_species$sim+1

write_csv(sim_output_species, 
          path = here("data/chromevol-simulations-outputs-species.csv"))
```

## Null model simulations

Create null simulations by randomly assigning chromosome numbers to the tips. The numbers come from the data and sampling is done without replacement.

```{r}
# Read in observed data
mydata <- read_csv(here("data/chromosome-data.csv"))

# Remove Rieppeleon
mydata <- 
  mydata %>% 
  filter(genus != "Rieppeleon")

# Select one representative per species
mydata_species <-
  mydata %>%
  mutate(dup = duplicated(mydata$Binomial_Marcello)) %>%
  filter(dup != TRUE)

# Rename the haploidn column
# Select only tip names, genus and observed haploidn
mydata_species <-
  mydata_species %>%
  rename(obs_haploidn = haploidn) %>%
  dplyr::select(tips, obs_haploidn, genus)
```

Run simulations
```{r}
sims_output_null_species <- data.frame(sim_haploidn = NULL, tips = NULL, sim = NULL)
  
for (j in 1:1000){
  newdata <- data.frame(sim_haploidn = sample(mydata_species$obs_haploidn, 
                                              length(mydata_species$tips)),
                        tips = mydata_species$tips, 
                        sim = rep(j, length(mydata_species$tips)))

  sims_output_null_species <- rbind(sims_output_null_species, newdata)
}

write_csv(sims_output_null_species, 
          path = here("data/null-simulations-outputs-species.csv"))
```