---
title: "Taxa and species pairs analyses"
author: "Natalie Cooper"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(ape)
library(phangorn)
```

Read in the data and the tree
```{r, message = FALSE, echo = FALSE}
mydata <- read_csv(here("data/chromosome-data.csv"))
mytree <- read.tree(here("data/Bayesian-tree.tre"))
```

Take a look at the dataset and tree
```{r}
#str(mytree)
#head(mydata)
```

Initially we use the phylogeny as is, and use OTUs which are often within the same species. As a test, we correlated the difference in chromosome number with phylogenetic distance.

1. Look at all pairs of taxa at the tips. 

This is the simplest solution in many ways, but does introduce a large amount of pseudoreplication as each tip taxon appears a huge number of times.
```{r}
# Get species pairs
all.pairs <- expand.grid(tip1 = mydata$tips,
	                       tip2 = mydata$tips)

# Combine with the chromosome data...
all.pairs  <- left_join(all.pairs, mydata, by = c("tip1" = "tips"))
all.pairs  <- left_join(all.pairs, mydata, by = c("tip2" = "tips"))

# Find cophenetic distances for whole tree
phylo_matrix <- cophenetic.phylo(mytree)

# Flatten matrix
phylo.dist.all <- 
  data.frame(expand.grid(tip1 = dimnames(phylo_matrix)[[1]],
	                       tip2 = dimnames(phylo_matrix)[[2]]),
             phylodist = c(phylo_matrix))

# Add to pairs data
# warns about adding vectors and factors - this is fine here
pairs.data <- 
  left_join(phylo.dist.all, all.pairs, by = c("tip1","tip2")) %>% 
  filter(phylodist > 0)
```

Now work out the differences in chromosome numbers
```{r}
pairs.data2 <-
  pairs.data %>%
  mutate(chromon_diff = abs(chromon.x - chromon.y))
```

And plot...
```{r}
# hex plot  
ggplot(pairs.data2, aes(x = phylodist, y = chromon_diff)) +
  geom_hex(alpha = 0.8, bins = 20) +
  theme_bw(base_size = 14) +
  scale_fill_viridis_c(trans = "log10", option = "plasma") +
  theme(strip.background = element_rect(fill = "white")) +
  xlab("phylogenetic distance (Ma)") +
  ylab("difference in chromosome number (n)")

#ggsave(filename = here("outputs/OTUs/taxon-pairs-distances.png"), height = 5)
```