---
title: "Ecological correlates analyses - figures"
author: "Natalie Cooper"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparation

```{r, message = FALSE, echo = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(ape)
library(broom)
library(patchwork)
```

Read in the data and the tree
```{r, message = FALSE, echo = FALSE}
mydata <- read_csv(here("data/chromosome-data-species.csv"))
mytree <- read.tree(here("data/Bayesian-tree-species.tre"))

# Remove outgroup
mydata <- mydata %>%
  filter(genus != "Leiolepis")
```

Modify some variables to make plotting/analyses easier

Remove complex levels from Substrate and relevel so multiple is last in the plots
```{r}
spdata <-
  mydata %>%
  mutate(substrate2 = case_when(Substrate == "Arboreal&Terrestrial" ~ "Multiple",
                                Substrate == "Arboreal&Saxicolous&Terrestrial" ~ "Multiple",
                                TRUE ~ as.character(Substrate))) %>%
  mutate(substrate2 = factor(substrate2, levels = c("Arboreal", "Terrestrial", "Multiple")))
```

## Plotting relationships 

```{r}
# Create list of colours to match earlier plots
mycolours <- c("#208EA3", "#FD817D", "#4178BC", "#EA4E9D", 
               "#EECC16", "#E8384F","#E37CFF","#37A862", "#A4C61A", "#7A71F6")

#"#E8384F" = RED
#"#208EA3" = AQUA
#"#FD817D" = ORANGE
#"#4178BC" = BLUE
#"#FDAE33" = YELLOW ORANGE
#"#7A71F6" = INDIGO
#"#EECC16" = YELLOW
#"#AA71FF" = PURPLE
#"#A4C61A" = YELLOW GREEN
#"#E37CFF" = MAGENTA
#"#62BB35" = GREEN
#"#EA4E9D" = HOT PINK
#"#37A862" = BLUE GREEN
```

Make series of plots for the appendix

### Genus
```{r}
genus <-
  ggplot(mydata, aes(y = haploidn, x = genus, colour = genus)) +
  geom_jitter(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6, face = "italic"),
        legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("Genus") +
  scale_colour_manual(values = mycolours)
```

### Realm
```{r}
realm <-
  ggplot(spdata, aes(y = haploidn, x = Realm, colour = genus)) +
  geom_jitter(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6),
        legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("biogeographic realm") +
  scale_colour_manual(values = mycolours)
```

### Absolute latitude
```{r}
latitude <- 
  ggplot(spdata, aes(y = haploidn, x = abs(Latitude), colour = genus)) +
  geom_point(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("absolute latitude (DD)") +
  scale_colour_manual(values = mycolours)
```

### Max SVL
```{r}
svl <-
  ggplot(spdata, aes(y = haploidn, x = maxSVL, colour = genus)) +
  geom_point(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("max SVL (mm)") +
  scale_colour_manual(values = mycolours)
```

### Substrate
```{r}
substrate <- 
  ggplot(data = subset(spdata, !is.na(substrate2)), 
       aes(y = haploidn, x = substrate2, colour = genus)) +
  geom_jitter(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6),
        legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("substrate") +
  scale_colour_manual(values = mycolours)
```

### Reproductive mode
```{r}
repro <-
  ggplot(data = subset(spdata, !is.na(Repromode)), 
       aes(y = haploidn, x = Repromode, colour = genus)) +
  geom_jitter(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6),
        legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("reproductive mode") +
  scale_colour_manual(values = mycolours)
```

### Min Clutch size
```{r}
minclutch <-
  ggplot(data = subset(spdata, !is.na(minClutchSize)), 
       aes(y = haploidn, x = minClutchSize, colour = genus)) +
  geom_point(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("min clutch size (eggs)") +
  scale_colour_manual(values = mycolours)
```

### Max Clutch size
```{r}
maxclutch <-
  ggplot(data = subset(spdata, !is.na(maxClutchSize)), 
       aes(y = haploidn, x = maxClutchSize, colour = genus)) +
  geom_point(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("max clutch size (eggs)") +
  scale_colour_manual(values = mycolours)
```

### Min breeding age
```{r}
minbreed <-
  ggplot(data = subset(spdata, !is.na(minBreedingAge)), 
       aes(y = haploidn, x = minBreedingAge, colour = genus)) +
  geom_point(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("min breeding age (months)") +
  scale_colour_manual(values = mycolours)
```

### Max Breeding Age
```{r}
maxbreed <-
  ggplot(data = subset(spdata, !is.na(maxBreedingAge)), 
       aes(y = haploidn, x = maxBreedingAge, colour = genus)) +
  geom_point(alpha = 0.6) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") +
  ylab("chromosomes (n)") +
  xlab("max breeding age (months)") +
  scale_colour_manual(values = mycolours)
```

And plot them all!
```{r}
(genus)/(realm + latitude) 

#ggsave(filename = here("outputs/ecology-genus-geo.png"), height = 5)   
   
(genus)/(svl + substrate + repro)

#ggsave(filename = here("outputs/ecology-genus-svl.png"), height = 5)   
  
(minclutch + maxclutch + minbreed + maxbreed)

#ggsave(filename = here("outputs/ecology-breeding.png"), height = 5)  
```