---
title: "Ecological correlates analyses - MCMCGLMM"
author: "Natalie Cooper"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparation

```{r, message = FALSE, echo = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(ape)
library(broom)
library(patchwork)
library(MCMCglmm)
library(coda)
library(phytools)
```

Read in the data and the tree
```{r, message = FALSE, echo = FALSE}
mydata <- read_csv(here("data/chromosome-data.csv"))
mytree <- read.tree(here("data/ML-tree.tre"))
mytree <- force.ultrametric(mytree)
```

Collapse data and tree to the species level using Meiri species, and remove the 5 species with no matching data
```{r}
spdata <-
  mydata %>%
  mutate(dup = duplicated(mydata$Binomial)) %>%
  filter(dup == FALSE) %>%
  filter(!is.na(Realm))

# Remove extra tips
remove <- setdiff(mytree$tip.label, spdata$tips)
sptree <- drop.tip(mytree, remove)

# Remove node labels
sptree$node.label <-NULL

# Check both have 80 species
#spdata
#sptree
```
Modify some variables to make plotting/analyses easier

Remove complex levels from Substrate and relevel so multiple is last in the plots
```{r}
spdata <-
  spdata %>%
  mutate(substrate2 = case_when(Substrate == "Arboreal&Terrestrial" ~ "Multiple",
                                Substrate == "Arboreal&Saxicolous&Terrestrial" ~ "Multiple",
                                TRUE ~ as.character(Substrate))) %>%
  mutate(substrate2 = factor(substrate2, levels = c("Arboreal", "Terrestrial", "Multiple"))) %>%
  as.data.frame()
```

Rescale numeric variables so they are mean centered and scaled to unit variance
```{r}
spdata <-
  spdata %>%
  mutate(maxSVL_Z = scale(log(spdata$maxSVL))) %>%
  mutate(lat_Z = scale(abs(spdata$Latitude))) %>%
  mutate(minClutchSize_Z = scale(spdata$minClutchSize)) %>%
  mutate(maxClutchSize_Z = scale(spdata$maxClutchSize)) %>%
  mutate(minBreedingAge_Z = scale(log(spdata$minBreedingAge))) %>%
  mutate(maxBreedingAge_Z = scale(log(spdata$maxBreedingAge)))
```

Create subsets of data and tree for analyses without full 80 species
```{r}
# Substrate
spdata_substrate <- 
  spdata %>%
  filter(!is.na(Substrate))

sptree_substrate <- drop.tip(sptree, setdiff(sptree$tip.label, spdata_substrate$tips))

# Reproductive mode
spdata_repro <- 
  spdata %>%
  filter(!is.na(Repromode))

sptree_repro <- drop.tip(sptree, setdiff(sptree$tip.label, spdata_repro$tips))

# Clutch size
spdata_clutch <- 
  spdata %>%
  filter(!is.na(minClutchSize))

sptree_clutch <- drop.tip(sptree, setdiff(sptree$tip.label, spdata_clutch$tips))

# Breeding age
spdata_breed <- 
  spdata %>%
  filter(!is.na(minBreedingAge))

sptree_breed <- drop.tip(sptree, setdiff(sptree$tip.label, spdata_breed$tips))
```

## Fit MCMCglmm models

Set up priors for modelling
```{r}
# Priors
prior <- list(R = list(V = 1, nu = 0.002), 
               G = list(G1 = list(V = 1, nu = 1, 
                                  alpha.mu = 0, alpha.V = 25^2)))

# Number of iterations
nitt <- 1E06

# Sampling interval
thin <- 1E03

# Burnin
burnin <- 1E05
```

Make trees - these will vary depending on variable
```{r}
# Inverse tree vcv matrix
# For 80 species
inv <- inverseA(sptree, scale = FALSE)$Ainv

# Substrate
inv_substrate <- inverseA(sptree_substrate, scale = FALSE)$Ainv

# Reproductive mode
inv_repro <- inverseA(sptree_repro, scale = FALSE)$Ainv

# Clutch size
inv_clutch <- inverseA(sptree_clutch, scale = FALSE)$Ainv

# Breeding age
inv_breed <- inverseA(sptree_breed, scale = FALSE)$Ainv
```

Fit models! These are Bayesian so might take a while to run...

### SVL
```{r}
svl_model <- MCMCglmm(haploidn ~ maxSVL_Z, 
                   data = spdata, random = ~ tips,
                   ginverse = list(tips = inv), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(svl_model$Sol)  

# For random effects
plot(svl_model$VCV)

# Look for autocorrelation
autocorr(svl_model$VCV)

# Calculate ESS
effectiveSize(svl_model$Sol[, 1:svl_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(svl_model)$solutions

# Get lambda
lambda <- svl_model$VCV[,'tips']/
          (svl_model$VCV[,'tips'] + svl_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

max_SVL <- summary(svl_model)$solutions[2,]
```

### Realm
```{r}
realm_model <- MCMCglmm(haploidn ~ Realm, 
                   data = spdata, random = ~ tips,
                   ginverse = list(tips = inv), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(realm_model$Sol)  

# For random effects
plot(realm_model$VCV)

# Look for autocorrelation
autocorr(realm_model$VCV)

# Calculate ESS
effectiveSize(realm_model$Sol[, 1:realm_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(realm_model)$solutions

# Get lambda
lambda <- realm_model$VCV[,'tips']/
          (realm_model$VCV[,'tips'] + realm_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

RealmMadagascar <- summary(realm_model)$solutions[2,]
RealmOriental <- summary(realm_model)$solutions[3,] 
RealmPalearctic <- summary(realm_model)$solutions[4,] 
```

### Latitude
```{r}
lat_model <- MCMCglmm(haploidn ~ lat_Z, 
                   data = spdata, random = ~ tips,
                   ginverse = list(tips = inv), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(lat_model$Sol)  

# For random effects
plot(lat_model$VCV)

# Look for autocorrelation
autocorr(lat_model$VCV)

# Calculate ESS
effectiveSize(lat_model$Sol[, 1:lat_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(lat_model)$solutions

# Get lambda
lambda <- lat_model$VCV[,'tips']/
          (lat_model$VCV[,'tips'] + lat_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

Latitude <- summary(lat_model)$solutions[2,]
```

### Substrate
```{r}
sub_model <- MCMCglmm(haploidn ~ substrate2, 
                   data = spdata_substrate, random = ~ tips,
                   ginverse = list(tips = inv_substrate), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(sub_model$Sol)  

# For random effects
plot(sub_model$VCV)

# Look for autocorrelation
autocorr(sub_model$VCV)

# Calculate ESS
effectiveSize(sub_model$Sol[, 1:sub_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(sub_model)$solutions

# Get lambda
lambda <- sub_model$VCV[,'tips']/
          (sub_model$VCV[,'tips'] + sub_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

SubstrateTerrestrial <- summary(sub_model)$solutions[2,]
SubstrateMultiple <- summary(sub_model)$solutions[3,]
```

### Reproductive mode
```{r}
repro_model <- MCMCglmm(haploidn ~ Repromode, 
                   data = spdata_repro, random = ~ tips,
                   ginverse = list(tips = inv_repro), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(repro_model$Sol)  

# For random effects
plot(repro_model$VCV)

# Look for autocorrelation
autocorr(repro_model$VCV)

# Calculate ESS
effectiveSize(repro_model$Sol[, 1:repro_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(repro_model)$solutions

# Get lambda
lambda <- repro_model$VCV[,'tips']/
          (repro_model$VCV[,'tips'] + repro_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

ReproductiveMode <- summary(repro_model)$solutions[2,]
```

### Min clutch size
```{r}
clutch1_model <- MCMCglmm(haploidn ~ minClutchSize_Z, 
                   data = spdata_clutch, random = ~ tips,
                   ginverse = list(tips = inv_clutch), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(clutch1_model$Sol)  

# For random effects
plot(clutch1_model$VCV)

# Look for autocorrelation
autocorr(clutch1_model$VCV)

# Calculate ESS
effectiveSize(clutch1_model$Sol[, 1:clutch1_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(clutch1_model)$solutions

# Get lambda
lambda <- clutch1_model$VCV[,'tips']/
          (clutch1_model$VCV[,'tips'] + clutch1_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

min_ClutchSize <- summary(clutch1_model)$solutions[2,]
```

### Max clutch size
```{r}
clutch2_model <- MCMCglmm(haploidn ~ maxClutchSize_Z, 
                   data = spdata_clutch, random = ~ tips,
                   ginverse = list(tips = inv_clutch), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(clutch2_model$Sol)  

# For random effects
plot(clutch2_model$VCV)

# Look for autocorrelation
autocorr(clutch2_model$VCV)

# Calculate ESS
effectiveSize(clutch2_model$Sol[, 1:clutch2_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(clutch2_model)$solutions

# Get lambda
lambda <- clutch2_model$VCV[,'tips']/
          (clutch2_model$VCV[,'tips'] + clutch2_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

max_ClutchSize <- summary(clutch2_model)$solutions[2,]
```

### Min breeding age
```{r}
breed1_model <- MCMCglmm(haploidn ~ minBreedingAge_Z, 
                   data = spdata_breed, random = ~ tips,
                   ginverse = list(tips = inv_breed), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(breed1_model$Sol)  

# For random effects
plot(breed1_model$VCV)

# Look for autocorrelation
autocorr(breed1_model$VCV)

# Calculate ESS
effectiveSize(breed1_model$Sol[, 1:breed1_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(breed1_model)$solutions

# Get lambda
lambda <- breed1_model$VCV[,'tips']/
          (breed1_model$VCV[,'tips'] + breed1_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

min_BreedingAge <- summary(breed1_model)$solutions[2,]
```

### Max breeding age
```{r}
breed2_model <- MCMCglmm(haploidn ~ maxBreedingAge_Z, 
                   data = spdata_breed, random = ~ tips,
                   ginverse = list(tips = inv_breed), family = "poisson", prior = prior,
                   nitt = nitt, thin = thin, burnin = burnin, pl = TRUE, verbose = FALSE)

# Plot model diagnostics for MCMCglmm
# Fixed effects
plot(breed2_model$Sol)  

# For random effects
plot(breed2_model$VCV)

# Look for autocorrelation
autocorr(breed2_model$VCV)

# Calculate ESS
effectiveSize(breed2_model$Sol[, 1:breed2_model$Fixed$nfl, 
                                 drop = FALSE])[[1]]

# Look at summary of fixed effect results
summary(breed2_model)$solutions

# Get lambda
lambda <- breed2_model$VCV[,'tips']/
          (breed2_model$VCV[,'tips'] + breed2_model$VCV[,'units'])

# Posterior mode of lambda
posterior.mode(lambda)

# HPD interval
HPDinterval(lambda)

max_BreedingAge <- summary(breed2_model)$solutions[2,]
```

Combine results
```{r}
results <- cbind(max_SVL, SubstrateTerrestrial, SubstrateMultiple, 
                 ReproductiveMode, min_ClutchSize, 
                 max_ClutchSize, min_BreedingAge, 
                 max_BreedingAge, RealmMadagascar, Latitude)
results <- as.data.frame(t(results))

results <-
  results %>%
  mutate(variable = rownames(results)) %>%
  select(variable, everything())

#write_csv(results, path = here("outputs/models-ecology.csv"))
```

Plot
```{r}
# Read in results
results <- read_csv(here("outputs/models-ecology.csv"))

# Relevel
results <- 
  results %>%
  mutate(variable = factor(variable, levels = c("max_SVL", "SubstrateTerrestrial",
                                                "SubstrateMultiple",
                                                "ReproductiveMode", "min_ClutchSize",
                                                "max_ClutchSize", "min_BreedingAge", 
                                                "max_BreedingAge", "RealmMadagascar",
                                                "Latitude")))

ggplot(results, aes(x = (variable), y = post.mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = `l-95% CI`, ymax = `u-95% CI`), width = 0.1) +
  theme_bw(base_size = 14) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ylab("posterior mean slope estimate") +
  xlab("") +
  coord_flip()

#ggsave(filename = here("outputs/mcmcglmm-figure.png"), height = 5)
```


